Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-13 Manuscript/SS_GAM.tex
Replace all ^\tp with *

\author[a]{Boyi Guo*}
\author[b]{Byron C. Jaeger}
\author[a]{AKM Fazlur Rahman}
\author[a]{D. Leann Long}
\author[a]{Nengjun Yi}

@@

\corres{Nengjun Yi, Department of Biostatistics, University of Alabama
at Birmingham, Birmingham, USA. \email{nyi@uab.edu}}

@@

\abstract{There are many proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional
(p\textgreater\textgreater n) problem using group penalties and priors.
These models rarely smooth the nonlinear effect properly because of the
sparse regularization. It leads to excess shrinkage on coefficients and
inaccurate prediction. Besides, most of these GAMs consider an
``all-in-all-out'' approach for function selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address the shortcomings, using Markov chain Monte
Carlo (MCMC) algorithms for model fitting creates a new challenge,
scalability. Hence, we propose Bayesian hierarchical generalized
additive models as a solution. For smooth and sparse solutions, the
proposed models employ bi-part spike-and-slab spline priors, including
mixture double exponential and mixture normal priors. Two
optimization-based algorithms, EM-Coordinate Descent and EM-Iterative
Weighted Least Squares, are developed for expedited computation.
Simulation studies and metabolomics data analyses demonstrate improved
performance against state-of-art models, mgcv, COSSO and sparse Bayesian
GAM. The software implementation of the proposed models is freely
available via an R package BHAM.}

@@

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{\text{pen}}
\newcommand{\pr}{\text{Pr}}

@@

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. As these methods focus on addressing the sparsity,
it is very likely to overly penalize the basis function coefficients and
produce inaccurate predictions and curve interpolation, particularly
when complex signals are assumed and large number of knots are used.
\citep{Scheipl2013} In addition, these methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for bi-level selection. Scheipl et al.
\citep{Scheipl2012} proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model framework.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties re also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions, also
know as the null space, from the nonlinear space. We then impose a new
bi-part spike-and-slab spline prior on the smoothing functions for
bi-level selection such that the linear and nonlinear spaces of
smoothing functions can be selected separately. The prior setup
encourages a flexible solution, rending one of three possibilities: no
effect, only linear effect, or nonlinear effect of a predictor. In
addition, two fast computing optimization-based algorithms are developed
for high and ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies.
Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothing penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function, or least squared and likelihood function
in the penalized framework. However, the direct integration of smoothing
penalties into sparsity penalties is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix is most often symmetric and positive semi-definite for
univariate smoothing functions, we apply eigendecomposition on the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smoothing function. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\)
containing the eigenvectors as columns in cthe orresponding order. That
is \(\boldsymbol{U}^0\) contains the eigenvectors with zero eigenvalues
for the null space and \(\boldsymbol{U}^{\text{pen}}\) contains the
eigenvectors (as columns) for the non-zero eigenvalues, i.e.~the
penalized space. We multiply the basis function matrix
\(\boldsymbol{X}\) with the orthonormal matrix \(\boldsymbol{U}\) for
the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) can receive uniform
penalty on each of its dimensions. With slight abuse of the notation, we
denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + {\boldsymbol{\beta}_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

The re-parameterization step sets up the foundation of the proposed
model and provides three-fold benefits. First of all, the
re-parameterization integrates the smoothing penalty into the design
matrix, and encourages models to properly smooth the nonlinear function
in addition to the sparse penalty for function selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear from the nonlinear space, improving the feasibility of bi-level
function selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^\text{pen}\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summarize,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{\text{spike}}(\cdot)\) for weak
signals and a flat slab density \(f_{\text{slab}}(\cdot)\) for strong
signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma\) that indicates whether the variable
\(x\) is included in the model. There are various spike-and-slab priors
depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab priors receive is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.
\end{equation} The SSL also mitigates the problem of EMVS where the weak
signals are not shrink to zero, and hence is prefered in
high-dimensional data analysis. We notice that Bai \citep{Bai2021} is
the first to apply spike-and-slab lasso prior in the GAM framework,
where the densities of the spike and slab components take the group
lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^\text{pen}\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the null space coefficients and on the
penalized space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1) \nonumber \\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator variable
and vectors, indicating if the model includes the linear effect and the
nonlinear effect of the \(j\)th variable respectively. \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
add a more restrictive assumption on the priors, assuming that one
indicator variable decides the inclusion of both the linear effect and
nonlinear effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\).
This converges to the SB-GAM \citep{Bai2021}. Conversely, it is also
possible to relax the assumption such that each coefficient \(\beta\) to
have its own latent indicator \(\gamma\), but at the cost of
complicating the bi-level function selection. This reduces the proposed
prior to the classic spike-and-slab lasso prior.

@@

\begin{itemize}
\tightlist
\item
  E-step: to calculate the expectation of the likelihood function
  conditioning on some ``missing'' data
\item
  M-step: to maximize the spectated likelihood function to calculate the
  parameter of interest
\end{itemize}

>>>>

\author[a]{Boyi Guo*}
\author[b]{Byron C. Jaeger}
\author[a]{AKM Fazlur Rahman}
\author[a]{D. Leann Long}
\author[a]{Nengjun Yi*}

@@

\corres{Boyi Guo and Nengjun Yi, Department of Biostatistics, University
of Alabama at Birmingham, Birmingham, USA. \email{boyiguo1@uab.edu} and
\email{nyi@uab.edu}}

@@

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional
(p\textgreater\textgreater n) problem using group penalties and priors.
These models rarely smooth the nonlinear effect properly because of the
sparse regularization. It leads to excess shrinkage on coefficients and
inaccurate prediction. Besides, most of these GAMs consider an
``all-in-all-out'' approach for function selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address the shortcomings, using Markov chain Monte
Carlo (MCMC) algorithms for model fitting creates a new challenge,
scalability. Hence, we propose Bayesian hierarchical generalized
additive models as a solution. For smooth and sparse solutions, the
proposed models employ bi-part spike-and-slab spline priors, including
mixture double exponential and mixture normal priors. Two
optimization-based algorithms, EM-Coordinate Descent and EM-Iterative
Weighted Least Squares, are developed for expedited computation.
Simulation studies and metabolomics data analyses demonstrate improved
performance against state-of-art models, mgcv, COSSO and sparse Bayesian
GAM. The software implementation of the proposed models is freely
available via an R package BHAM.}

@@

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{*}
\newcommand{\pr}{\text{Pr}}

@@

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. As these methods focus on addressing the sparsity,
it is very likely to overly penalize the basis function coefficients and
produce inaccurate predictions and curve interpolation, particularly
when complex signals are assumed and large number of knots are used.
\citep{Scheipl2013} In addition, these methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for bi-level selection. Scheipl et al.
\citep{Scheipl2012} proposed a spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model framework.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties were also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions, also
know as the null space, from the nonlinear space. We then impose a new
bi-part spike-and-slab spline prior on the smoothing functions for
bi-level selection such that the linear and nonlinear spaces of
smoothing functions can be selected separately. The prior setup
encourages a flexible solution, rending one of three possibilities: no
effect, only linear effect, or nonlinear effect of a predictor. In
addition, two fast computing optimization-based algorithms are developed
for high and ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via Monte Carlo simulation studies.
Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(y|X\beta, \phi) = \prod\limits^n_{i=1}f(y_i|X_i,\phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothing penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function, or least squared and likelihood function
in the penalized framework. However, the direct integration of smoothing
penalties into sparsity penalties is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix is most often symmetric and positive semi-definite for
univariate smoothing functions, we apply eigendecomposition on the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the linear space of the
smoothing function. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order. That
is \(\boldsymbol{U}^0\) contains the eigenvectors with zero eigenvalues
for the linear space and \(\boldsymbol{U}^{*}\) contains the
eigenvectors (as columns) for the non-zero eigenvalues, i.e.~the
non-linear space. We multiply the basis function matrix
\(\boldsymbol{X}\) with the orthonormal matrix \(\boldsymbol{U}\) for
the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive uniform penalty on
each of its dimensions. With slight abuse of the notation, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right].
\end{equation}

The re-parameterization step sets up the foundation of the proposed
model and provides three-fold benefits. First of all, the
re-parameterization integrates the smoothing penalty into the design
matrix, and encourages models to properly smooth the nonlinear function
in addition to the sparse penalty for function selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear from the nonlinear space, improving the feasibility of bi-level
function selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^*\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summarize,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{\text{spike}}(\cdot)\) for weak
signals and a flat slab density \(f_{\text{slab}}(\cdot)\) for strong
signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma \in \{0,1\}\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab priors receive is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
a double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, (1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the linear space coefficients and on the
nonlinear space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\sim DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and \(\gamma^*_{j}\in \{0,1\}^{K_j}\)
are latent indicator variable and vectors, indicating if the model
includes the linear effect and the nonlinear effect of the \(j\)th
variable respectively. \(s_0\) and \(s_1\) are scale parameters,
assuming \(0 < s_0 < s_1\) and given. These scale parameters \(s_0\) and
\(s_1\) can be treated as tuning parameters and optimized via
cross-validation. A discussion of how to choose the scale parameters
comes in Section \ref{sec:tune}. To note, this prior differs from
previous group spike-and-slab lasso priors \citep{Tang2018, Tang2019},
where the \(\boldsymbol{\beta}^0_j\) and \(\boldsymbol{\beta}^*_j\)
shares the same indicator variables \(\gamma_j^0\), \(\gamma_j^*\)
respectively. It is possible to add a more restrictive assumption on the
priors, assuming that one indicator variable decides the inclusion of
both the linear effect and nonlinear effect,
i.e.~\(\gamma_j = \gamma^0_j = \gamma^*_j\). This converges to the
SB-GAM \citep{Bai2021}. Conversely, it is also possible to relax the
assumption such that each coefficient \(\beta\) to have its own latent
indicator \(\gamma\), but at the cost of complicating the bi-level
function selection. This reduces the proposed prior to the classic
spike-and-slab lasso prior.

@@

\begin{itemize}
\tightlist
\item
  E-step: to calculate the expectation of the likelihood function
  conditioning on some ``missing'' data
\item
  M-step: to maximize the expected likelihood function to calculate the
  parameter of interest
\end{itemize}
